---
title: "Anions and Cations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anions and Cations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


Balances of anions and cations are used throughout bcgwcat in order to assess
various water chemistry metrics.

In particular they are used to calculate

 - **Charge Balances** `charge_balance()`
 - **Water Types** `water_type()`
 - **Piper plots** `piper_plot()`
 - **Stiff plots** `stiff_plot()`

Generally speaking, charge balances are used to assess whether the water chemistry
data is valid. For water types, piper plots and stiff plots, samples are only 
retained if the charge balance is <=10%. 

The exact types of anions and cations used depend on the calculation.

```{r}
#| echo: false
#| results: "asis"
ions <- data.frame(
  ions = c("Cl", "SO4", "HCO3", "CO3", "F", "NO3", "NO2", "Meas_Alk",
           "Ca", "Mg", "Na", "K", "Al_diss", "Cu_diss", "Fe_diss", "Mn_diss", "Zn_diss", "NH4",
           "pH_lab"),
  type = c(rep("anions", 8), rep("cations", 11))) |>
  dplyr::mutate(
    `charge balance` = dplyr::if_else(ions %in% c(
      "Cl", "SO4", "F", "NO3", "NO2", "Meas_Alk",
      "Ca", "Mg", "Na", "K", "Al_diss", "Cu_diss", "Fe_diss", "Mn_diss", "Zn_diss", "NH4", "pH_lab"),
      "X", ""),
    `water type` =  dplyr::if_else(ions %in% c("Cl", "SO4", "HCO3", "Ca", "Mg", "Na", "K"), "X", ""),
    `water type` = dplyr::if_else(ions == "Meas_Alk", "*", `water type`),
    `piper plots` = dplyr::if_else(ions %in% c("Cl", "SO4", "HCO3", "CO3", "Ca", "Mg", "Na", "K"),  "X", ""),
    `piper plots` = dplyr::if_else(ions == "Meas_Alk", "*", `piper plots`),
    `stiff plots` = dplyr::if_else(ions %in% c("Cl", "SO4", "HCO3", "Ca", "Mg", "Na"),  "X", ""),
    `stiff plots` = dplyr::if_else(ions == "Meas_Alk", "*", `stiff plots`),
  )

ions |>
  dplyr::group_by(type) |>
  gt::gt() |>
  gt::cols_align("center", columns = -"ions") |>
  gt::tab_style(gt::cell_text(weight = "bold", align = "center", transform = "capitalize"),
                locations = gt::cells_column_labels(-"ions")) |>
  gt::tab_style(gt::cell_text(weight = "bold", transform = "capitalize"),
                locations = list(gt::cells_row_groups(), gt::cells_column_labels("ions"))) |>
  gt::tab_footnote("* Used if HCO3 is missing")
```


### Charge balances

Charge balances are calculated based on a formula from ALS formula.

Although EMS data contains charge balances, as well as the anion and cation sums
used to calculate these values, potential changes in workflows over the years
have made it difficult to ascertain exactly how charge balances were calculated
in older samples. This resulted in discrepancies between EMS and locally
calculated charge balances. Therefore for consistency, we calculate charge
balances for all samples using the ALS formula below.

**Note:** This means that the original EMS charge balances, anion sums, and 
cation sums, have been **omitted**.

One difference between our calculation and that of ALS, is that we use more
significant digits when calculating MEQ.

Charge balances use following measures in MEQ

- **Anions** Cl, SO4, F, NO3, NO2, Meas_Alk
- **Cations** Ca, Mg, Na, K, Al_diss, Cu_diss, Fe_diss, Mn_diss, Zn_diss, NH4,
  and pH_lab

And are calculated as follows

$\textrm{Anion sum} = Cl + SO_4 + F + NO_3 + NO_2 + Meas\_Alk$

$$
\textrm{Cation sum} = Ca + Mg + Na + K + Al_{diss} +
             Cu_{diss} + Fe_{diss} + Mn_{diss} + Zn_{diss} + NH_4 +
             (10 ^{-\textrm{pH_lab}}) * 1000
$$

$\textrm{Charge balance} = | 100 \times (\textrm{Cation Sum} - \textrm{Anion sum}) / (\textrm{Cation Sum} + \textrm{Anion Sum}) |$

Missing values are ignored (i.e. generally treated as 0). However, if all
values for cations or anions are missing the charge balance is `NA`.

This means that Charge Balances will be invalid if there are too many missing
measures.

### Water Types

Water types identify the major cations and anions in a water sample. 
In bcgwcat, we calculate water types using the following ions:

- **Anions** Cl, SO4, HCO3
- **Cations** Ca, Mg, and Na

Water type is only calculated for samples with valid charge balances.
Ions are ranked by proportion MEQ, all greater than 10% are listed in descending
order, cations first. If HCO3 is missing, we use Meas Alk as a replacement
(indicated by an '*' on HCO3 in the water type). Otherwise, missing ions are
ignored (i.e. treated as 0). The column `missing_ion` indicates whether there
is a problem with the water type, such that it is missing a cation or anion
(i.e. is all cations or all anions).

For example with this data
```{r}
#| echo: false
d <- data.frame(Sample_Date = "2022-01-01", SampleID = "999990-01", StationID = 000,
                Cl_meq = 0.0226, SO4_meq = 0.0208, HCO3_meq = 1.54, Meas_Alk_meq = 1.7,
                Ca_meq = 0.187, Mg_meq = 0.490, Na_meq = 0.465, 
                charge_balance = 0.5) |>
  dplyr::as_tibble()
gt::gt(d)
```

We would get the following water type.
```{r}
#| echo: false
gt::gt(bcgwcat::water_type(d)[, c("water_type", "missing_ion")])
```
